### 2013-07-07 doc/dissertation/document/chapters/parallel_mc/parallel_mc.tex
finished parallel chapter draft

To motivate this problem, consider the square domain presented in
Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
Carlo particle transport problem. If the domain were decomposed into 9
subdomains as shown, each of those subdomains and their associated
data (i.e. cross sections) could be owned by a different parallel
process in the computation. In Figure~\ref{fig:ddmc_example}, the
tracks of 3 particles that are born in the center subdomain are
shown. In this example, particle A is first transported from the
center domain to the domain directly to the left. Before the
scattering event in the new domain may be processed, particle A must
be communicated between between the two parallel processes that own
those domains. For any given particle in the transport simulation,
this communication event may hardly occur during the lifetime of the
particle (particle B), or may occur many times (particle C) depending
on the problem parameters and the random path in phase space taken by
the particle. Scalable parallel algorithms for domain decomposed Monte
Carlo are those that handle this domain-to-domain communication of
particles effectively and balance it with the on-process computations
for particle interactions, tallies, response calculations, and other
simulation requirements.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.}}
  \label{fig:ddmc_example}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposed Neumann-Ulam Algorithm\ }
\label{sec:asynchronous_algorithm}
In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel algorithm as
implemented in production implicit Monte Carlo codes
\citep{brunner_efficient_2009}. We will adapt their algorithm and
directly apply it to a parallel formulation of the Neumann-Ulam
method. Direct analogs can be derived from their works by noting that
the primary difference between solving a linear transport system with
Monte Carlo methods and traditional fixed source Monte Carlo transport
problems is the content of the Markov chains that are generated. The
transitions represented by these chains are bound by probabilities and
weights and are initiated by the sampling of a discrete source. In the
context of transport problems, those transitions represent events such
as particle scattering and absorption with probabilities that are
determined by physical data in the form of cross sections.

For stochastic matrix inversion, those transitions represent moving
between the equations of the linear system (and therefore the
components of phase space which they represent) and their
probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver.

In this section Brunner and Brantley's fully asynchronous algorithm is
presented along with its application to the Neumann-Ulam method. For
considerably more detail, the algorithm presented can be found in
\citep{brunner_efficient_2009} and was effectively implemented
verbatim for this work. In their work they identify two data sets that
are required to be communicated: the sharing of particles that are
transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Both of these
communication sequences will be addressed at a high level along with
how the Monte Carlo data structures they require are constructed in
parallel.

@@

Compared to the serial construction of the Neumann-Ulam decomposition,
data for all states that is required to be on process must be
collected in parallel. For the pure domain decomposition case, given
an input parallel decomposition for the Neumann-Ulam decomposition,
for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
adjacency graph of a single state in the neutron diffusion matrix for
the model problem presented in Appendix~\ref{chap:diffusion_problem}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
  \end{center}
  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
    \textit{Adjacent boundary states are collected by traversing the
      graph one step for all local boundary states and collecting the
      equations for those states that are not local. For a given
      transition in the random walk for the diffusion problem, a
      history at mesh point $(i,j)$ may transition to all adjacent
      mesh points including itself, corresponding to a global state
      transition from $m$ to $n$ in the linear system.}}
  \label{fig:diffusion_graph}
\end{figure}

@@

\subsection{The Algorithm}
\label{subsec:parallel_mc_algorithm}

@@

Successful execution of this algorithm requires construction of the
parallel Neumann-Ulam decomposition as described in the preceding
section. This data is used to begin
Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
collected list of neighboring processors generated while building the
Neumann-Ulam decomposition to setup the set of asynchronous messages
required for domain-to-domain communication. Again, consider this
communication patter for the 9 subdomain example given by
Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each domain
has two neighbors with which they will communicate parallel histories
and this communication goes both ways as represented by the adjacent
arrows in the figure. For each set of neighbors, a non-blocking send
and receive is required with data buffers allocated with a
user-defined size prepared for incoming and outgoing histories. This
nonblocking structure is critical to the performance of the algorithm
in that it permits local history transport to receive while new work
to do is being collected. When a given process is ready to do more
work, it can check these data buffers for incoming histories requiring
further transport. In this way there is a maximum amount of overlap
between communication and transport of histories.

@@

To generate overlap, we use the identical algorithm but in this case
we perform as many graph traversals as the specified amount of
overlap. For example, consider the analytic relations derived in
\S~\ref{sec:analytic_framework} for the length of random walks in a
Neumann-Ulam method and the amount of leakage from a domain
considering a neutron diffusion problem. From these relations we might
determine that for our particular problem, if we can grow the domain
by 10 additional discrete state transitions, we can reduce the number
of histories that must be communicated by a significant fraction. We
determine the information we must gather from neighboring domains to
build the overlap by traversing the graph of the matrix outward from
the boundary 10 times. At each step, we find the data in the adjacent
domain that we require and make that data local, incrementing the size
of the overlap. Once the overlap has been gathered, one extra graph
traversal on the boundary states is performed to get the new
neighboring states and their owning process.

@@

Once complete, the tallies for each individual computation must be
gathered to the original set in order to build the correction vector
within an MCSA sequence. To do this, we perform two parallel vector
operations. First, overlap not only creates additional states in the
local transport domain but also additional states in the local tally
vector such that any events that occur in the local domain may be
tallied in a local vector. Because of this, the tally vector in a
single set will share local states with other domains and a parallel
sum reduction operation is required to transform the tally into the
original Neumann-Ulam parallel decomposition. Second, the tallies in
individual sets must be combined and applied to the primary set in the
problem that shares the processor space with the original linear
problem. For this operation, we employ the concept of blocks as a
subdivision of parallel space complementary to the concept of sets.

Consider the schematic representation of the final parallel reduction
presented in Figure~\ref{fig:msod_tally_reduction}. For this
operation, the tally vectors in each set must be summed together and
applied to the primary set. As each set is an exact replica off the
others, the parallel decomposition of the tally vector is the same
within all sets. We can take advantage of this by building a processor
space that encapsulates each identical domain in each set. We can then
use this processor space for the reduction. For the schematic in
Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
set creates a single block, giving 9 total blocks for this example
problem. The processor space the encapsulates each of the upper-left
domains is then used for the parallel reduction operation. This
reduction will happen in a mutually exclusive manner for each of the
blocks in the system. One benefit of this approach is that this
reduction occurs only over a subset of the processors in the problem,
providing better scalability than a reduction operation over all
processors in the problem.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions to the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, the solution at each grid point
in the system will be compared to the solution at the same grid point
in the serial reference case with the minimum and maximum absolute
differences of these values reported.

GMRES and conjugate gradient solvers from the Belos package were used
on 4 cores as a means of both assuring the Krylov solvers produce the
same results on multiple cores and as a means of providing an
acceptable range of minimum and maximum values in the comparison. For
MCSA, a range of MSOD parameters will be used such that problems with
multiple sets and multiple blocks may be tested along with
overlap. When multiple sets were leveraged, a single history was used
to compute the MCSA correction for every DOF in the problem. For all
solvers, the flux was converged to a tolerance of \sn{1}{-8} and
therefore if the comparisons agree within this tolerance the solutions
will be deemed
correct. Table~\ref{tab:parallel_verification_parameters} gives the
parameters for the verification problem and
Table~\ref{tab:parallel_mcsa_verification} gives the results of the
calculations. Solver parameters used for MCSA are given in
Table~\ref{tab:parallel_mcsa_parameters}.

@@

From the data we see that for all solvers, the parallel results agree
within the solution tolerance of the reference serial results for all
DOF in the system. In addition, we do not notice any systematic
differences in the results as a function of number of blocks, number
of sets, or amount of overlap. In addition, the parallel GMRES results
have maximum absolute differences of the same magnitude as all MCSA
results and a larger minimum absolute difference than all MCSA results
by three orders of magnitude. Based on these results, MCSA parallelized
with an MSOD, and in particular the implementation used here,
generates the correct results with respect to the serial conjugate
gradient reference computation.

@@

How effectively the presented algorithms parallelize MCSA will be
determined by several performance measures. For this work, these
measures come directly from Keyes' primer on the scalability of domain
decomposed methods and specifically his discussion on how to measure
their parallel performance \citep{keyes_how_1999}. All of these
metrics are derived from two measured quantities: the number of
iterations required for convergence and the total wall time required
for convergence. Each of these quantities are then measured in varying
parameter studies by changing the number of processes in a simulation
and by varying the local or global problem size. Using the notation of
Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
where the efficiency is a function of the global problem size, $N$,
and the number of processes in the calculation, $P$. There are several
types of efficiencies that will be useful in our analysis of
scalability with different definitions depending on the type of
parametric scaling study performed. For all efficiencies, a value of 1
is considered perfect while a value of 0 signals the worst possible
performance. Often, as was observed in this work, it is the case that
efficiencies above 1 are computed in certain scaling studies. For the
data presented in the following sections, an explanation of each
observed case of super-unitary efficiencies will be provided.

@@

In addition to strong scaling, in certain cases it is useful to
consider the \textit{weak scaling} of an iterative method. In this
case, the ratio $N/P$ is fixed such that the local problem size
remains the same for all values of $P$. In this case, the objective is
to maintain a constant wall time while both $N$ and $P$ grow as a
fixed ratio. In this case, we define the absolute efficiency as:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed. Often, weak scaling
is of interest to the nuclear engineer due to the fact that the
computational resources available are fixed and may not be able to
contain the entire global problem of interest. In this case, an
algorithm with good weak scaling performance permits all of the
computational resources available to be used effectively to allow for
the highest fidelity problem possible to be solved. In addition, for
the resources available, the memory to processor ratio on a
distributed machine is often fixed, giving an exercise in fixed $N/P$
to solve the highest fidelity problem possible. Therefore, good weak
scaling should not be viewed as an enhancement of time to solution for
the engineer but rather permitting a higher fidelity problem to be
solved within a given time constraint. For problems where increasing
fidelity increases the stiffness of the system, there is often a
trade-off between the absolute efficiencies computed for a given
method and the degrading algorithmic efficiencies that arise when
increasing the fidelity. Both time and hardware constraints should be
considered when preparing a parallel calculation.

@@

For the scaling results presented here, all parallel Monte Carlo
results were computed using the Monte Carlo Linear
Solvers\footnote{MCLS is an open source library and can be downloaded
  at \url{https://github.com/sslattery/MCLS}} (MCLS) library generated
specifically for this work while the conjugate gradient and GMRES
Krylov solvers used are part of the Belos linear solver package within
the Trilinos framework \citep{heroux_overview_2005}. In general, the
Belos package produces solutions to the neutron diffusion problem
nearly two orders of magnitude faster than the MCLS library for serial
computations using the conjugate gradient solver. Therefore, when we
directly compare these two codes, the MCLS parallel efficiencies will
appear artificially inflated relative to the Belos efficiencies due to
the fact that the MCLS ratio of work to communication will be much
larger than that for the Belos calculations. For every scaling study
presented, the calculations were performed 3 times with the average
values reported for wall time and used for the efficiency
computations. In addition, it should be noted that in practice,
symmetric positive-definite problems like the neutron diffusion
problem solved here would never be solved using GMRES. Instead, a
method such as conjugate gradient would typically be used due to the
enhanced performance. We use GMRES here to simply as a means of
comparison to MCSA performance.

@@

For the scaling studies presented in this work, only the CPU
components of Titan specified in Table~\ref{tab:titan_hardware} were
utilized. This is due to the fact that the implementations of the
parallel MCSA algorithm utilized code written only with the Message
Passing Interface (MPI) and therefore could not take advantage of the
GPU accelerators. Even considering just the CPU portion of Titan,
nearly 300,000 CPUs are available to test the parallel performance of
the MCSA algorithm at high levels of concurrency.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Case}& 
      \multicolumn{1}{c}{Iterations}&
      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
      1 & 22 & 1.0 \\
      2-split & 22 & 1.0 \\
      2-replicated & 16 & 1.375 \\
      4-split & 22 & 1.0 \\
      4-replicated & 13 & 1.692 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Strong scaling algorithmic efficiency using
      multiple sets.}}
  \label{tab:ms_strong_alg_eff}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Case}& 
      \multicolumn{1}{c}{Iterations}&
      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
      1 & 22 & 1.0 \\
      2-split & 22 & 1.0 \\
      2-replicated & 16 & 1.375 \\
      4-split & 22 & 1.0 \\
      4-replicated & 13 & 1.692 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
      sets.}}
  \label{tab:ms_weak_alg_eff}
\end{table}

>>>>

To motivate this problem, consider the square domain presented in
Figure~\ref{fig:ddmc_example} and on this domain we imagine a Monte
Carlo particle transport problem. If the domain were decomposed into 9
subdomains as shown, each of those subdomains and their associated
data (i.e. cross sections) could be owned by a different parallel
process in the computation. In Figure~\ref{fig:ddmc_example}, the
tracks of 3 particles that are born in the center subdomain are
shown. In this example, particle A is first transported from the
center domain to the domain directly to the left. Before the
scattering event in the new domain may be processed, particle A must
be communicated between between the two parallel processes that own
those domains. For any given particle in the transport simulation,
this communication event may hardly occur during the lifetime of the
particle (particle B), or may occur many times (particle C) depending
on the problem parameters and the random path in phase space taken by
the particle. Scalable parallel algorithms for domain decomposed Monte
Carlo are those that handle this domain-to-domain communication of
particles effectively and balance it with the on-process computations
for particle interactions, tallies, response calculations, and other
simulation requirements.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{
      \input{chapters/parallel_mc/ddmc_example.pdftex_t} }
  \end{center}
  \caption{\textbf{Domain decomposed Monte Carlo transport example
      illustrating how domain decomposition requires parallel
      communication of particles.} \textit{If a particle crosses a
      domain boundary it must be communicated to the next domain in
      its path.}}
  \label{fig:ddmc_example}
\end{figure}

@@

%%---------------------------------------------------------------------------%%
\section{Domain Decomposed Neumann-Ulam Algorithm\ }
\label{sec:asynchronous_algorithm}
In the context of radiation transport, in 2009 Brunner and colleagues
provided a fully asynchronous domain decomposed parallel algorithm as
implemented in production implicit Monte Carlo codes
\citep{brunner_efficient_2009}. We will adapt their algorithm and
directly apply it to a parallel formulation of the Neumann-Ulam
method. Direct analogs can be derived from their works by noting that
the primary difference between solving a linear transport system with
Monte Carlo methods and traditional fixed source Monte Carlo transport
problems is the content of the Markov chains that are generated.

The transitions represented by these chains are bound by probabilities
and weights and are initiated by the sampling of a discrete source. In
the context of transport problems, those transitions represent events
such as particle scattering and absorption with probabilities that are
determined by physical data in the form of cross sections. For
stochastic matrix inversion, those transitions represent moving
between the equations of the linear system (and therefore the
components of phase space which they represent) and their
probabilities are defined by the coefficients of those
equations. Ultimately, we tally the contributions to generate
expectation values in the desired states as we progress through the
chains. Therefore, parallel methods for Monte Carlo radiation
transport can be abstracted and we can use those concepts that apply
to matrix inversion methods as an initial means of developing a
parallel Neumann-Ulam-type solver.

In this section Brunner and Brantley's fully asynchronous algorithm,
which was effectively implemented verbatim for this work, is presented
along with its application to the Neumann-Ulam method. For
considerably more detail, the algorithm presented can be found in
\citep{brunner_efficient_2009}. In their work they identify two data
sets that are required to be communicated: the sharing of particles
that are transported from one domain to another and therefore from one
processor to another and a global communication that signals if
particle transport has been completed on all processors. Both of these
communication sequences will be addressed at a high level along with
how the Monte Carlo data structures they require are constructed in
parallel.

@@

Compared to the serial construction of the Neumann-Ulam decomposition,
data for all states that is required to be on process must be
collected in parallel. For the pure domain decomposition case, given
an input parallel decomposition for the Neumann-Ulam decomposition,
for all local states $m$ we require access to $P_{mn}$ and $W_{mn}$
(or $P^T_{mn}$ and $W^T_{mn}$) for the adjoint method for all possible
states $n$. Given by Figure~\ref{fig:diffusion_graph}, consider the
adjacency graph of a single state in the neutron diffusion matrix for
the model problem presented in Appendix~\ref{chap:diffusion_problem}.
\begin{figure}[t!]
  \begin{center}
    \scalebox{1.5}{ \input{chapters/parallel_mc/stencil_graph.pdftex_t} }
  \end{center}
  \caption{\textbf{Neutron diffusion equation state adjacency graph.}
    \textit{The structure of the graph comes from the discretization
      of the Laplacian operator describing the diffusion
      physics. Adjacent boundary states are collected by traversing
      the graph one step for all local boundary states and collecting
      the equations for those states that are not local. For a given
      transition in the random walk for the diffusion problem, a
      history at mesh point $(i,j)$ may transition to all adjacent
      mesh points including itself, corresponding to a global state
      transition from $m$ to $n$ in the linear system.}}
  \label{fig:diffusion_graph}
\end{figure}

@@

\subsection{Domain Decomposed Algorithm}
\label{subsec:parallel_mc_algorithm}

@@

Successful execution of this algorithm requires construction of the
parallel Neumann-Ulam decomposition as described in the preceding
section. This data is used to begin
Algorithm~\ref{alg:parallel_mc_algorithm} where lines 1-5 use the
collected list of neighboring processors generated while building the
Neumann-Ulam decomposition to setup the set of asynchronous messages
required for domain-to-domain communication. Again, consider this
communication pattern for the 9 subdomain example given by
Figure~\ref{fig:nearest_neighbor_comm}. In this pattern, each boundary
domain has two neighbors and the center domain four neighbors with
which they will communicate parallel histories and this communication
goes both ways as represented by the adjacent arrows in the
figure. For each set of neighbors, a non-blocking send and receive is
required with data buffers allocated with a user-defined size prepared
for incoming and outgoing histories. This nonblocking structure is
critical to the performance of the algorithm in that it permits local
history transport to continue while new histories to transport are
being collected in the buffers. When a given process is ready to do
more work, it can check these data buffers for incoming histories
requiring further transport. In this way there is a maximum amount of
overlap between communication and transport of histories.

@@

To generate overlap, we use the identical algorithm but in this case
we perform as many graph level traversals as the specified amount of
overlap. For example, consider the analytic relations derived in
\S~\ref{sec:analytic_framework} for the length of random walks in a
Neumann-Ulam method and the amount of leakage from a domain
considering a neutron diffusion problem. From these relations we might
determine that, for our particular problem, if we can grow the domain
by 10 additional discrete state transitions, we can reduce the number
of histories that must be communicated by a significant fraction. We
determine the information we must gather from neighboring domains to
build the overlap by traversing the graph of the matrix outward from
the boundary 10 levels. At each step, we find the data in the adjacent
domain that we require and make that data local, incrementing the size
of the overlap. Once the overlap has been gathered, one extra graph
traversal on the boundary states is performed to get the new
neighboring states and their owning processes.

@@

Once Monte Carlo transport is complete complete, the tallies for each
individual computation must be gathered to the original set in order
to build the correction vector within an MCSA sequence. To do this, we
perform two parallel vector operations. First, overlap not only
creates additional states in the local transport domain but also
additional states in the local tally vector such that any events that
occur in the local domain may be tallied in a local vector. Because of
this, the tally vector in a single set will share local states with
other domains and a parallel sum reduction operation is required to
transform the tally into the original Neumann-Ulam parallel
decomposition. Second, the tallies in individual sets must be combined
and applied to the primary set in the problem that shares the
processor space with the original linear problem. For this operation,
we employ the concept of blocks as a subdivision of parallel space
complementary to the concept of sets.

Consider the schematic representation of the final parallel reduction
presented in Figure~\ref{fig:msod_tally_reduction}. For this
operation, the tally vectors in each set must be summed together and
applied to the primary set. As each set is an exact replica of the
others, the parallel decomposition of the tally vector is the same
within all sets. We can take advantage of this by building a processor
space that encapsulates each identical domain in each set. We can then
use this processor space for the reduction. For the schematic in
Figure~\ref{fig:msod_tally_reduction}, the upper-left domain in each
set creates a single block, giving 9 total blocks for this example
problem. The processor space that encapsulates each of the upper-left
domains is then used for the parallel reduction operation. This
reduction will happen in a mutually exclusive manner for each of the
blocks in the system. One benefit of this approach is that this
reduction occurs only over a subset of the processors in the problem,
providing better scalability than a reduction operation over all
processors in the problem.

@@

%%---------------------------------------------------------------------------%%
\section{Parallel MCSA Verification\ }
\label{sec:parallel_verification}
Before we consider the parallel performance of the MCSA algorithm, we
must first verify that the algorithm generates the correct solution in
parallel with respect to reference solutions provided by production
Krylov methods. To verify correctness, we use solutions to the neutron
diffusion system presented in
Appendix~\ref{chap:diffusion_problem}. As the reference computation,
serial results will be used from the conjugate gradient solver in
Belos, a linear solvers package in the Trilinos scientific computing
libraries \citep{heroux_overview_2005}. In identical fashion to the
verification for the $SP_N$ equations performed in
\S~\ref{sec:spn_mcsa_verification}, we will solve the neutron
diffusion problem in parallel using multiple cores on a desktop
machine. For each parallel solution, then the solution at each grid
point in the system will be compared to the solution at the same grid
point in the serial reference case with the minimum and maximum
absolute differences of these values reported.

GMRES and conjugate gradient solvers from the Belos package were used
on 4 cores as a means of both assuring the Krylov solvers produce the
same results on multiple cores and as a means of providing an
acceptable range of minimum and maximum values in the comparison. For
MCSA, a range of MSOD parameters will be used such that problems with
multiple sets and multiple blocks may be tested along with
overlap. For all MCSA calculations, a single history was used for
every DOF in the problem to compute the correction vector. For all
solvers, the flux was converged to a tolerance of \sn{1}{-8} and
therefore if the comparisons agree within this tolerance the solutions
will be deemed
correct. Table~\ref{tab:parallel_verification_parameters} gives the
parameters for the verification problem and
Table~\ref{tab:parallel_mcsa_verification} gives the results of the
calculations. Solver parameters used for MCSA are given in
Table~\ref{tab:parallel_mcsa_parameters}.

@@

From the data we see that for all solvers, the parallel results agree
with the reference serial results within the solution tolerance for
all DOF in the system. In addition, we do not notice any systematic
differences in the results as a function of number of blocks, number
of sets, or amount of overlap. In addition, the parallel GMRES results
have maximum absolute differences of the same magnitude as all MCSA
results and a larger minimum absolute difference than all MCSA results
by three orders of magnitude. Based on these results, MCSA
parallelized with the MSOD algorithm, and in particular the
implementation used here, generates the correct results with respect
to the serial conjugate gradient reference computation.

@@

How effectively the presented algorithms parallelize MCSA will be
determined by several performance measures. For this work, these
measures will come directly from Keyes' primer on the scalability of
domain decomposed methods and specifically his discussion on how to
measure their parallel performance \citep{keyes_how_1999}. All of
these metrics are derived from two measured quantities: the number of
iterations required for convergence and the total wall time required
for convergence. Each of these quantities are then measured in varying
parameter studies by changing the number of processes in a simulation
and by varying the local or global problem size. Using the notation of
Keyes, we will define \textit{parallel efficiency} as $\eta(N,P)$
where the efficiency is a function of the global problem size, $N$,
and the number of processes in the calculation, $P$. There are several
types of efficiencies that will be useful in our analysis of
scalability with different definitions depending on the type of
parametric scaling study performed. For all efficiencies, a value of
100\% is considered perfect while a value of 0\% signals the worst
possible performance. Often, as was observed in this work, it is the
case that efficiencies above 100\% are computed in certain scaling
studies. For the data presented in the following sections, an
explanation of each observed case of super-unitary efficiencies will
be provided.

@@

In addition to strong scaling, in certain cases it is useful to
consider the \textit{weak scaling} of an iterative method. In this
case, the ratio $N/P$ is fixed such that the local problem size
remains the same for all values of $P$. In this case, the objective is
to maintain a constant wall time while both $N$ and $P$ grow as a
fixed ratio. In this case, we define the absolute efficiency as:
\begin{equation}
  \eta_{weak}(M|N,P|Q) = \frac{T(M,Q)}{T(N,P)}\:,
  \label{eq:weak_scaling_absolute}
\end{equation}
subject to the constraint $M/Q = N/P$. The scaling is perfect if the
runtime is static for all $P$ when $N/P$ is fixed.

Often, weak scaling is of interest to the nuclear engineer due to the
fact that the computational resources available are fixed and may not
be able to contain the entire global problem of interest. In this
case, an algorithm with good weak scaling performance permits all of
the computational resources available to be used effectively to allow
for the highest fidelity problem possible to be solved. In addition,
for the resources available, the memory to processor ratio on a
distributed machine is often fixed, giving an exercise in fixed $N/P$
to solve the highest fidelity problem possible. Therefore, good weak
scaling should not be viewed as an enhancement of time to solution for
the engineer but rather permitting a higher fidelity problem to be
solved within a given time constraint. For problems where increasing
fidelity increases the stiffness of the system, there is often a
trade-off between the absolute efficiencies computed for a given
method and the degrading algorithmic efficiencies that arise when
increasing the fidelity. Both time and hardware constraints should be
considered when preparing a parallel calculation.

@@

For the scaling results presented here, all parallel MCSA results were
computed using the Monte Carlo Linear Solvers\footnote{MCLS is an open
  source C++ library built on the Trilinos framework and can be
  downloaded at \url{https://github.com/sslattery/MCLS}} (MCLS)
library generated specifically for this work while the conjugate
gradient and GMRES Krylov solvers used are part of the Belos linear
solver package within the Trilinos framework
\citep{heroux_overview_2005}. In general, the Belos package produces
solutions to the neutron diffusion problem nearly two orders of
magnitude faster than the MCLS library for serial computations using
the conjugate gradient solver. Therefore, when we directly compare
these two codes, the MCLS parallel efficiencies will appear
artificially inflated relative to the Belos efficiencies due to the
fact that the MCLS ratio of work to communication will be much larger
than that for the Belos calculations. For every scaling study
presented, the calculations were performed 3 times with the average
values reported for wall time and those average values used for the
efficiency computations. In addition, it should be noted that in
practice, symmetric positive-definite problems like the neutron
diffusion problem solved here would never be solved using
GMRES. Instead, a method such as conjugate gradient would typically be
used due to the enhanced performance. We use GMRES here simply as a
means of providing an additional comparison for MCSA performance.

@@

For the scaling studies presented in this work, only the CPU
components of Titan specified in Table~\ref{tab:titan_hardware} were
utilized. This is due to the fact that the implementations of the
parallel MCSA algorithm in the MCLS library utilized code written only
with the Message Passing Interface (MPI) and therefore could not take
advantage of the GPU accelerators. Even considering just the CPU
portion of Titan, nearly 300,000 CPUs are available to test the
parallel performance of the MCSA algorithm at high levels of
concurrency.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Case}& 
      \multicolumn{1}{c}{Iterations}&
      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
      1 & 22 & 1.0 \\
      2-split & 22 & 1.0 \\
      2-replicated & 16 & 1.375 \\
      4-split & 22 & 1.0 \\
      4-replicated & 13 & 1.692 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Strong scaling algorithmic efficiency using
      multiple sets.} \textit{Replicating histories improves
      algorithmic efficiency by converging in fewer iterations.}}
  \label{tab:ms_strong_alg_eff}
\end{table}

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Case}& 
      \multicolumn{1}{c}{Iterations}&
      \multicolumn{1}{c}{$\eta_{alg}$} \\\hline
      1 & 22 & 1.0 \\
      2-split & 22 & 1.0 \\
      2-replicated & 16 & 1.375 \\
      4-split & 22 & 1.0 \\
      4-replicated & 13 & 1.692 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Weak scaling algorithmic efficiency using multiple
      sets.} \textit{Replicating histories improves algorithmic
      efficiency by converging in fewer iterations.}}
  \label{tab:ms_weak_alg_eff}
\end{table}
